
# Overview

## Project description

This analysis aims to characterize approximately 25 years of shoreline profile transect data, provided by NANOOS (Northwest Association of Networked Ocean Observing Systems).

This repository focuses specifically on Westport Light State Park, profiles 16 and 17 drawn from raw data consisting of approximately 4500 individual profiles across 54 coastal sites in Washington and Oregon.

The profiles are in x y z format (Easting Northing elevation).The naming convention is prof_X_ssYY.out where prof is short for profile, X is the profile number, ss is a season code (e.g. f = fall) and YY is the year, in the format of the last two digits (e.g. "98" is 1998", "00" is 2000, "08" is 2008, etc).

## Data discrepancies

Profiles that are null are removed. 
A small portion of the profiles had non-conforming filenames: "BigE06", "beachface", etc.
These files do not represent a large percentage of the files and have not been included in the analysis.

Profiles 42 - 47 are in Oregon and have been excluded from most of the analysis.

*Geographically, profiles do not proceed sequentially.*

## Metadata

The vertical datum is NAVD88.
The horizontal is WA State Plane South.
All of the units are in meters.

**The profiles are converted to feet/year for this Westport analysis. They are in meters/year for the WA State analysis.**

# ```{r knit_setup, include=FALSE}
# ## Script Setup
# source("scripts/src/kableExtrafunction.R")
# 
# knitr::opts_chunk$set(
#   echo = FALSE,
#   warning = FALSE,
#   message = FALSE
#   )
# options(knitr.kable.NA = "") 
# ```

```{r load_packages, include=FALSE, echo=FALSE, results='hide'}
## Load required packages
source("scripts/src/load_packages.R")
```

```{r examine_profiles, echo=FALSE, include=FALSE}
source("scripts/examine_profiles.R")
```

## List of profiles and their geographic locations

```{r profile_location}
profiles.w.location %>% 
  knitr::kable(
    caption = "Profiles with Corresponding Geographic Location",
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    align = "l"
    ) %>%
  kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15",
      fixed_thead = TRUE
    )
profiles.w.location %>%
  filter(profile %in% c(16, 17))
```

# Individual profiles, single year visualization

Below is a rendering of a single profile for a single year.

This is a direct plot of the x y z coordinates to understand the general idea of the data format.
It is *not* a geographic representation of the data.

```{r single_profile_visualization, echo=FALSE}
## Modify profile.pattern to select a given profile
profile.pattern <- "prof_16"

## Modify year.pattern to select a given year
year.pattern <- c("99")

## Render the chosen profile in 3D
source("scripts/single_profile_3D.R")

suppressWarnings(single.profile.plot)
```

# Individual profiles, all available years visualization

Below is a rendering of a single profile.
The orange horizontal plane represents the approximate mean high high water (MHHW) mark, as a proxy for understanding how the transect data is shifting over time.

```{r complete_profile_visualization, echo=FALSE}
profile.pattern <- "prof_17"

source("scripts/complete_profile_plane.R")

suppressWarnings(print(complete.profile.plot))
```

## Times series representation

It can also be helpful to see the above plot spread over time, to provide a clearer picture of profile migration.

```{r complete_profile_timeseries, echo=FALSE}
profile.pattern <- "prof_16"

source("scripts/complete_profile_timeseries.R")

suppressWarnings(print(profile.timeseries.fig))

```

# The "BasePoint" reference line and Euclidean distance as an accretion/erosion proxy

In order to quantify shoreline change, an arbitrary "BasePoint" reference line, approximately parallel to the shoreline, was drawn at a reasonable distance from the profiles so as to be unaffected by temporal change.

The NANOOS profiles are perpendicular to this BasePoint, allowing for a reference against which to calculate spatial change.

To visualize this concept, the images below display 1) a simplified selection of profiles (plotted in black) compared to the BasePoint (plotted in red) and 2) an individual profile (plotted in black) and its BasePoint (plotted in red).

```{r profiles_BasePoint, echo=FALSE}
source("scripts/profiles_and_BasePoint.R")
print(all.basepoint.plot)
print(single.basepoint.plot)
```


Using Euclidean distance, migrations of the transects over time as compared to the BasePoint act as proxies for accretion and erosion. 

The midpoint of each transect is calculated to determine a single comparative point using the midpoint formula pictured below.

$$\left(\frac{x_1+x_2}{2}, \frac{y_1+y_2}{2}\right)$$

The Euclidean distance between the midpoint and the fixed BasePoint is calculated for each transect and each year, resulting in an increasing or decreasing Euclidean distance. This distances correlates to an accreting or eroding transect.

$$d=\sqrt{(x_2-x_1)^2 + (y_2-y_1)^2}$$

|                               |           |
|-------------------------------|-----------|
| Increasing Euclidean distance | Accretion |
| Decreasing Euclidean distance | Erosion   |

# Determining significance in accretion and erosion classifications for profiles. 

Apply the midpoint formula and distance formula to each transect/BasePoint pair, and plot the movement of the midpoint over time. This produces an x:y, distance:time relationship to which a linear model can be fitted. Once the linear model has been produced, the slope and p values can be extracted and used as respective indicators of transect characterization and significance. 

If the slope of a given distance:time relationship is positive, this indicates the distance between the transect midpoint and the BasePoint is increasing and the shoreline is accreting. If the slope is negative, the distance is decreasing and the shoreline is eroding. Linear models with a pvalue < 0.05 are considered significant.

```{r midpoint_euclidean, echo=FALSE}
profile.pattern <- "prof" 
source("scripts/midpoint_euclidean_distance.R")
midpoint.euc.dist.plot
```



# Assign status to each profile according to linear regression equation details.

Below is a table with the relevant results of each profile's linear regression over time, including pvalues (less than 0.05 indicates a significant relationship), rsquared values (indicitive of how good the fit of the regression is) and the Pearson correlation (how closely are time and euclidean distance, or x and y, related).

```{r linear_parameters, echo=FALSE}
results.table %>% 
  knitr::kable(
    caption = "Linear equation parameters",
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    align = "l"
    ) %>%
  kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header", "scale_down"),
      stripe_color = "gray!15",
      fixed_thead = TRUE
    )
results.table
```

# Comparison of results to Washington Coastal Erosion Hazard Assessment (WCEHA)

The WCEHA (<https://waecy.maps.arcgis.com/apps/View/index.html?appid=389d0a3ce642485db912d4a416a56e25>) is a second database that can be used as a comparison to the results from the NANOOS profiles.

```{r WCEHA_validation, echo=FALSE}
source("scripts/WCEHA_validation.R")

all.data <- all.data %>% 
  knitr::kable(
    caption = "Comparison with WCEHA data",
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    align = "l"
    ) %>%
  kableExtra::kable_styling(
      position = "left",
      latex_options = c("scale_down", "repeat_header"),
      stripe_color = "gray!15",
      fixed_thead = TRUE
    ) %>%
  row_spec(which(all.data$conflict == "flag"), bold = T, color = "white",
           background = "red")
all.data

```

## Some notes on comparisons between the NANOOS analysis and the WCEHA data

-   Profiles 16 and 17 (Westport Light North and South)

    -   Both profiles' NANOOS classification of "accretion" conflicts with the WCEHA classifcation of "erosion". It's true that Westport North is eroding overall, but WCEHA contains data from 2006 - 2019, and NANOOS contains data from 1997 - 2022. NANOOS incorporates the geomorphologically impactful Army Corp nourishment more holistically. Both of the Westport profiles were "non significant" accretion, and speaks to the volatility of shorelines on the peninsula.
    
    
# Quartile Points and Change Rates

Using the midpoint of each transect selects a single point as a proxy for erosion through time.
This can obscure the inherent variation in the collection of profiles.

The profiles have roughly linear relationships due to the geographic layout: in general, a change in x (moving along the Easting plane) will result in a *roughly* equal change in y (the Northing plane).
Therefore, a linear regression can be applied to the x and y variables of each profile to obtain a line proxy rather than a single point.

Below are images of a single profile and year to see the applied linear regression. In the first image, a linear regression is overlaid onto a single profile, and in the second, the linear quartiles of the regression are added. Note that in the second image, the red "BasePoint" has been included to visualize distance. 

These quartiles are labeled "minimum", "quartile1", "median", "quartile3", and "maximum" in accordance with standard boxplot parameters, although they are calculated using *only distance* and not boxplot delineations.

```{r single_linear_quartiles16, echo=FALSE}
profile.pattern <- "prof_16"
year.pattern <- "00"

source("scripts/single_linear_quartiles.R")
single.linear.plot
single.quartiles.plot
```

```{r single_linear_quartiles17, echo=FALSE}
profile.pattern <- "prof_17"
year.pattern <- "00"

source("scripts/single_linear_quartiles.R")
single.linear.plot
single.quartiles.plot
```

Characterizing the profiles with more points allows for a more detailed look at erosion and accretion rates.

Each quartile's Euclidean movement is tracked over time, using the same calculations as the midpoint proxy used above. The rate of change is then calculated for each quartile of each profile, using a standard rate of change calculation between each year:

$$\frac{EucDist_2 - EucDist_1}{EucDist_1} \times 100\%$$
EucDist2 is the more recent year, and EucDist1 is the older year. This is calculated for each set of years from 1997 - 2022. 
For each profile, the average rate of change (ROC) between all quartiles is taken. 

```{r rates_of_change, echo=FALSE}
profile.pattern <- "prof_17"
source("scripts/rates_of_change.R")
profile.ROC.plot
annualized.median.plot
```
